---
title: "Brownian Notes"
author: "SQ"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
source("utilities.R")

```

# Brownian motion
a literal random motion, think things like random molecule movement or particle dispersion in air or water, but also conceptual, like for finance or econ things, queuing networks in operations research

Properties!:
Stationary increments, only depends on the length of the time period, meaning Bs+t - Bs has the same distribution as Bt, meaning it's probs are the same, and any Bt has a normal distribution N(0,t)
Independent increments too, separate spaces of time not affected by others
SBM, by the way, is Standard Brownian Motion

Example:
For 0 < s < t, find the distribution of Bs + Bt
well, since s is less than t, a Bs is inside our Bt, so se can translate Bs + Bt to 2Bs + (Bt-Bs),
and since these are now independent increments, they can have separate probabilities calculated!
also, the sum of indep normal rv is normal, so Bs + Bt is normal with mean E[Bs + Bt] = E[Bs] + E[Bt] = 0
and variance Var[2Bs + (Bt − Bs)] = Var[2Bs] + Var[Bt − Bs] = 4Var[Bs] + Var[Bt−s]
and since the variance of those independent distributions are s and t respectively, because of N(0,t), also
= 4s + (t − s) = 3s + t

```{r}
#EX: Particle position follows SBM, if its at position 1 at time t=2, what is the probability that its posiiton
# is at most 3 at time t=5? This tells us we want P{B5 ⩽ 3 | B2 = 1}, so a conditional prob, first off, we can
# write P{B5 − B2 ⩽ 3 − B2 | B2 = 1} as an equal, you remove the time from both. Because of the conditional, we
# know that B2 is 1, so we can remove it from the 3 part of the equation, and we know that 
# P{B5 − B2 ⩽ 2 | B2 = 1} = P{B5 − B2 ⩽ 2} is INDEPENDENT OF B2, because its outside its range, independent 
# increments property. And now, all you need is P{B3 ⩽ 2}, which we can get by just use
# pnorm for the cdf <= going (target, distribution mean, distribution sd (SO REMEMBER TO sqrt() THE t value))
pnorm(2, 0, sqrt(3)) #0.8758935


# EX2: Find the covariance of Bs and Bt, meaning we want Cov(Bs, Bt) = E[BsBt] − E[Bs]E[Bt]
# assuming s < t, we know Bt = Bs + (Bt − Bs), so we can plug that into E[BsBt] = E[Bs(Bs + (Bt − Bs))],
# then E[Bs^2 + Bs(Bt − Bs))] = E[Bs^2 ] + E[Bs]E[Bt − Bs] as these are all independent increments, and
# also, this just so happens to be the variance equation, so then Var[Bs] = s, meaning that the
# Cov(Bs,Bt) = min(s,t)


```
## Simulation of brownian motion

It calculates the position of the next point using a the previous point's position and a random normal distribution Xi as N(0,t/n) and adds them together, forming a recursive simulation method.
We then have a Z distribution of N(0,1), marking the top and bottom of our simulation

```{r}
n <- 1000
t <- 1
bm <- c(0, cumsum(rnorm(n, 0, sqrt(t/n))))
steps <- seq(0, t, length=n+1)
plot(steps, bm, type="l")
```

## Transformations of SBM
Reflection (with respect to the horizontal axis): -Bt where t>0
Translation: (Bs+t − Bs : t ⩾ 0)
Rescaling: (a^(−1/2)Bat : t ⩾ 0)
Inversion: process Xt = tB1/t, (inverting the time, observing the process at time 1/t!)

## Brownian Motion started at x
This refers to a process (Xt : t ⩾ 0) w/ Xt = x + Bt, where Bt belongs to an SBM

```{r}
# EX: Let {Xt : t ⩾ 0} be a Brownian motion process started at x = 3. Find P{X2 > 0}
# P{X2 > 0} = P{3 + B2 > 0} by properties, so then P{B2 > −3} = 1 − P{B2 ⩽ −3}
1-pnorm(-3, 0, sqrt(2)) #0.9830



```

## Combination with Markov
Brownian motion is a Markov process since it meets its requirements, and its a time-homogenous Markov process, meaning Bs+t given Bs is equal to Bt given B0, so P{Bs+t ⩽ y | Bs} = P{Bt ⩽ y}

## First Hitting times
Ta = min{t ⩾ 0 : Bt = a} is called the first hitting time to a

We have the property:
P{Bt > a | Ta < t} = P{Bt > 0} = 1/2, which then means, by Bayes theorem (slide 14)
1/2 = P{Bt > a}/P{Ta < t}
which means P{Ta < t} = 2P{Bt > a}

It will be hit with probability 1 in final time(?)

Distribution of the first hitting time:
P{Ta < t} = 2(1 − Φ(|a|/√t)) wherein Φ() is the cdf of N(0,1), a Z normal dist
By extension,
P{Ta < ∞} = 1
E[Ta] = ∞, and because of this, the variance is also infinite
meaning it will for sure get there, but it might take an infinite amount of time

```{r}
# EX: a particle moves according to brownian motion starting at x=1. After t=3 hours, it's at level 1.5,
# Find the probability the particle reaches level 2 sometime in the next hour
# First, we translate it, so for points t>=3, we can say the process started at x=1.5.
# Reaching 2 in the next hour starting at 1.5 is equal to an SBM reaching a=0.5 in its first hour
# So we want P{T0.5 ⩽ 1}, which we can get with the prior formula!
# P{Ta < t} = 2(1 − Φ(|a|/√t))
# P{T0.5 <= 1} = 2(1 − Φ(|0.5|/√1)), |0.5|/√1= 0.5, we put that in our target slot
2*(1-pnorm(0.5, 0, 1)) #0.617


```

## Distribution of the Maximum of Brownian Motion
Mt = max(Bs) within [0,t], and by properties, Mt and abs(Bt) have the same distribution.
Meaning P{Mt > a} == 2P{Z > a/√t} (slide 17)
This also tells us that P{Mt > a} = P{Ta < t}

```{r}
# EX: Errors increase over time in lab equipment, for how many years can a lab be guaranteed that there is at least 90% probability that all errors do not exceed 4 degrees? 
# We want the largest t subject to P{Mt ⩽ 4} ⩾ 0.9, prob maximum error is under 4 degrees,
# P{Mt ⩽ 4} = 1 − P{Mt > 4} = 1 − 2 P{Z > 4/√t} we do this conversion first, then we switch it around again
# = 1 − 2(1 − P{Z ⩽ 4/√t}) = (2P{Z ⩽ 4/√t}) − 1
# then 0.9 ⩽ 2 P{Z ⩽ 4/√t}−1  == 0.95 ⩽ P{Z ⩽ 4/√t}
qnorm(0.95,0,1) #gets you the Z probability it needs to beat
# meaning 4/sqrt(t) >= 1.645 and so t <= (4/1.645)^2
(4/1.645)^2 #5.91 years of usage

```

## Zeros of Brownian motion and arcsine laws 
The times when the Brownian motion crosses 0 are its zeroes (remember, generally brownian motion goes between -1 and 1, so its like the points where it crosses the midpoint.)
### Between zero
The prob that and SBM has a zero between times r and t is z(r,t), then
z(r,t) == (2/π) * arccos(sqrt(r/t)) == 1 - (2/π) * arcsin(sqrt(r/t))

### Last zero
The proof we have is that, given Lt is the last zero in (0, t), P{Lt ⩽ x} = (2/π) * arcsin(sqrt(x/t)), 0 < x < t
Meaning Lt/t is an Arcsin distribution
CDF: F(x) = (2/π) * arcsin(√x), 0 < x < 1
Density: f(x) = 1/(π * sqrt(x(1 − x))), 0 < x < 1

```{r}
# EX: Find the prob an SBM has at least 1 zero between times 3 and 5.
# This means we want z(3,5) == 1 - (2/π) * arcsin(sqrt(3/5))
1 - (2/pi)*asin(sqrt(3/5)) # 0.4359
# or (2/pi)*acos(sqrt(3/5))

# EX2: Find the prob that the last zero of sbm in (0,5) lies between 3 and 4
# We want P{3 < L5 ⩽ 4} = P{L5 ⩽ 4} − P{L5 ⩽ 3}
# that turns to (2/π) * arcsin(sqrt(4/5)) - (2/π) * arcsin(sqrt(3/5))
(2/pi)*asin(sqrt(4/5)) - (2/pi)*asin(sqrt(3/5)) # 0.1407


```

## Brownian Motion with drift
Since sbms are usually too simple, we describe Brownian motion with drift μ and variance σ^2 as
Xt = μt + σBt, t ⩾ 0
This process Xt still has independent and stationary increments
Xs+t − Xs ∼ Normal(μt, (σ^2)t)

```{r}
# EX: Find the prob that a brownian motion with drift μ=4 and variance = 0.25 takes values between
# 1 and 3 at time t=4. Since Xt ∼ Normal(0.6t, 0.25t), we do P{1 < Xt ⩽ 3} turning into Z standard forms
# P( (1-0.6t)/sqrt(0.25t) < Z <= (3-0.6t)/sqrt(0.25t)
t=4
pnorm((3-0.6 * t)/sqrt(0.25 * t)) - pnorm((1-0.6 * t)/sqrt(0.25 * t)) # either works, this is just straight up
pnorm(3, 0.6 * t, sqrt(0.25 * t)) - pnorm(1, 0.6 * t, sqrt(0.25 * t)) # 0.645

# EX2: Find the prob the home team wins a game given they are losing by two points at halftime
# using μ = 4.87, σ = 15.82. SInce Xt ∼ Normal(μt, (σ^2)t), and p(l,t) = P{X1−t > −l}
# then p(l = −2, t = 0.5) = P{X0.5 > 2}
t <- 0.50; mu <- 4.87; sigma <- 15.82
1 - pnorm(2, mu * (1-t), sigma * sqrt(1-t))

#oneliner
1 - pnorm(2, 4.87 * 0.5, 15.82 * 0.5) # 0.52

```

## Geometric Brownian Motion
used for modelling stock prices in the olden days
The process Gt, called geometric brownian motion, is defined by Gt = G0 * e^Xt
Since ln(G0) + Xt ∼ Normal(ln(G0) + μt, σ2t), Gt follows a logonormal distribution
This means it has certain properties
E[Gt] = G0 * e^(t(μ+(σ^2))/2))
Var[Gt] = (G0)^2 * e^(2t(μ+(σ^2))/2) * (e^(t(σ^2)) − 1)

```{r}
# EX: Assume a stock sells for 80eur and follows a geom brownian motion with yearly drift μ = 0.1
# and volatility σ = 0.5 find the prob that in 90 days, the price will rise to at least 100 eur
# If we make the t in Gt represent years, we can make 90 days as 0.25 years, so then P{G0.25 ⩾ 100}
# which is the same as P{G0 * e^X0.25 >= 100} == P{80 * e^X0.25 >= 100} == P{e^X0.25 >= 5/4} ==
# P{X0.25 >= ln(5/4)}
t <- 0.25; mu <- 0.1; sigma <- 0.5
1 - pnorm(log(5/4), mu * t, sigma * sqrt(t)) # wrong in slides, target is log(5/4)

#oneliner
1 - pnorm(log(5/4), 0.1 * 0.25, 0.5 * sqrt(0.25)) #pnorm uses target, mean and sd, so u get the sqrt instead of the square

```

Extending, You could represent daily stock changes as the multiplication of G0 and a bunch of Ys, where Ys = Gs/Gs-1, which are iid rvs with the same mean and vars as a Gt but without the G0 multiplication.

```{r}
# EX: Stock follows a geom brown with daily drift μ = 0.01 and volatility σ = 0.05. Find the prob
# the stock rises more than 2% from today to tomorrow. Find the prob it drops 2% from today to tomorrow.
# For the first part, we want P{G1/G0 > 1.02}, which is just P{e^X1 > 1.02} = P{X1 > ln(1.02)}, so
1 - pnorm(log(1.02), 0.01, 0.05) # 0.4223

# For the second part, we want P{e^X1 < 0.98} or P{X1 < ln(0.98}
pnorm(log(0.98), 0.01, 0.05) # 0.2729



```



We finished all the slides, btw






